---
title: "Session 2 - Self-Directed"
author: "Erik Steeb"
date: "2/10/2022"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Session 2 Overview

In this half of the workshop, we'll be working individually or in small groups on a provided dataset to apply what we learned in the first half of class. We'll be using a dataset from the [UC Irvine Machine Learning Repository] (https://archive.ics.uci.edu/ml/index.php). Specifically, we'll be looking at two .csv files with data about the physical characteristics and expert quality ratings of [white and red Vinho Verde] (https://archive.ics.uci.edu/ml/datasets/Wine+Quality).

If you are using the git repo, the files are in the datasets folder and can be loaded using the code below. If not, modify it to load the data yourself.

```{r load_data}
red_df <- read.csv('datasets/winequality-red.csv')
white_df <- read.csv('datasets/winequality-white.csv')


```


We'll also load the tidyverse set of packages so we're able to manipulate the data:

```{r}

library(tidyverse)

```

# Question: What factors have the biggest impact on percieved wine quality? Are these factors the same for white and red wine?